---
title: "13 EUI"
author:
  - name: "Lars Hinrichs"
    affiliation: "The University of Texas at Austin"
    email: "lh@utexas.edu"
    orcid: "0000-0003-3679-1927" 
  - name: "Stefan Dollinger"
    affiliation: "University of British Columbia"
    email: "stefan.dollinger@ubc.ca"
    orcid: "0000-0001-5582-5139" 
format:
  html:
    embed-resources: true
number-sections: true
toc: true
echo: false
---


```{r packages, include=FALSE}
rm(list=ls())
library(pacman)
p_load(rio, tidyverse, janitor, gt, tm)
```

```{r setup, message=FALSE, warning=FALSE, include=FALSE}

df <- import("data/df_cleaned_5.rds") %>% 
  as_tibble() %>% 
  filter(!region=="other")

```


## Variables

We now turn to `eui`. I has values on an integer scale from 0 to 15. 

```{r}
df <- df %>% 
  mutate(eui = as.integer(eui))
df %>% count(eui) %>% gt()
```
A quick plot of the distribution.

```{r fig-eui-density, fig.height=4.5, fig.width=8.5}

ncases <- nrow(df) %>% format(big.mark = ",")

ggplot(df, aes(x = eui)) +
  geom_density(fill = "blue", alpha = 0.5, bw = 0.5) +
  labs(title = "Density Plot of EUI",
       x = "EUI",
       y = "Density",
       caption = paste0("N = ", ncases)) +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0))


```




We are going to go ahead and bin the responses into groups: 

- Monolingual (0)
- Some non-English (1-2)
- Considerable non-English (3-15)

```{r}
df_eui <- df %>%
  mutate(eui_bin = case_when(
    eui == 0 ~ "Monolingual (0)",
    eui >= 1 & eui <= 2 ~ "Some non-English (1-2)",
    eui >= 3 & eui <= 15 ~ "Considerable non-English (3-15)",
    TRUE ~ NA_character_
  ),
  eui_bin = factor(eui_bin, levels = c("Monolingual (0)", "Some non-English (1-2)", "Considerable non-English (3-15)"))
  )

df_eui %>%  count(eui_bin) %>% print()

```

## Predicting `st_heard`

```{r fig-eui-bins-st-heard}
df_plot <- df_eui %>% 
  filter(!is.na(eui_bin),
         !is.na(st_heard))

ncases <- nrow(df_plot) %>% format(big.mark = ",")

p1 <- df_plot %>% 
ggplot(aes(x = eui_bin, fill = st_heard)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of st_heard as predicted by EUI",
       caption = bquote(italic(N) == .(ncases)),
       x = "EUI",
       y = "Count",
       fill = NULL) +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0))

p1 + scale_fill_manual(values = c("#0072B2","#E69F00", "#F0E442", "#CC79A7", "#D55E00")) 

```

```{r}
p1 + scale_fill_grey(start = 0.2, end = 0.8) 

```


```{r fig-eui-bins-st-heard-flip}
df_plot <- df_eui %>% 
  filter(!is.na(eui_bin),
         !is.na(st_heard))

ncases <- nrow(df_plot) %>% format(big.mark = ",")

p2 <- df_plot %>% 
ggplot(aes(x = st_heard, fill = eui_bin)) +
  geom_bar(position = "dodge", width = 0.7) +
  labs(title = "Distribution of EUI as predicted by st_heard",
       caption = bquote(italic(N) == .(ncases)),
       y = NULL,
       x = "st_heard",
       fill = NULL) +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0))

p2 + scale_fill_manual(values = c("#0072B2","#CC79A7", "#E69F00", "#F0E442",  "#D55E00")) 

```
  
```{r}
p2 + scale_fill_grey(start = 0.2, end = 0.8) 

```



```{r}

df_descr <- df_eui %>% 
  mutate(st_descr = str_trim(st_descr))

levs <- df_descr %>% pull(st_descr) %>% na.omit() %>% unique()

df_descr <- df_descr %>% 
  select(eui_bin, st_descr, region) %>%
  mutate(
    st_descr = factor(st_descr, levels = levs)
  ) %>% 
  na.omit()

df_descr %>% 
  count(eui_bin, st_descr) %>% print()
```

The data table (above) is what we want to visualize.

```{r fig-stdescr-eui, message=FALSE, fig.width=7, fig.height=4.5}


total_counts <- df_descr %>%
  group_by(eui_bin) %>%
  summarise(total_count = n())

data_percent <- df_descr %>%
  group_by(eui_bin, st_descr) %>%
  summarise(count = n()) %>%
  left_join(total_counts, by = "eui_bin") %>%
  mutate(percentage = count / total_count * 100)

# Calculate total counts for each eui_bin level
total_counts <- data_percent %>%
  group_by(eui_bin) %>%
  summarise(total_count = paste0("n = ", sum(count)))


# Create the plot
p5 <- ggplot(data_percent, aes(x = eui_bin, y = percentage, fill =  st_descr)) +
  geom_bar(stat = "identity", position = "stack", width = .6,
           show.legend = T) +
  geom_label(data = total_counts, aes(x = eui_bin, y = 60, label = total_count), vjust = -0.5, inherit.aes = FALSE) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Distribution of st_descr as predicted by EUI",
       x = NULL,
       y = "Percentage",
       fill = "St descr") +
  theme_minimal() +
  coord_flip() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0))


p5 + scale_fill_manual(values = c("#D55E00", "#0072B2","#CC79A7", "#E69F00", "#F0E442")) 

```

::: {.callout-tip}
## The lowdown
This distribution is not at all significant.
:::


## EUI and `cdn_way`

```{r}

levs <- df_eui %>% pull(cdn_way) %>% na.omit() %>% unique() %>% .[c(4,1,2,3)]


df_eui_way <- df_eui %>% 
  select(eui_bin, cdn_way, region) 

df_eui_way <- df_eui_way %>%
  mutate(
    cdn_way = factor(cdn_way, levels = levs)
  ) %>% 
    na.omit()

df_eui_way %>% 
  count(eui_bin, cdn_way)

```


```{r fig-cdnway-eui, message=FALSE, fig.width=7, fig.height=5}

# Calculate percentages and counts
data_percent <- df_eui_way %>%
  group_by(eui_bin, cdn_way) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Calculate total counts for each eui_bin level
total_counts <- data_percent %>%
  group_by(eui_bin) %>%
  summarise(total_count = paste0("n = ", sum(count)))

# Create the plot
p4 <- ggplot(data_percent, aes(x = eui_bin, y = percentage, fill = cdn_way)) +
  geom_bar(stat = "identity", position = "stack", width = .6) +
  geom_text(data = total_counts, aes(x = eui_bin, y = 100, label = total_count), vjust = -0.5, inherit.aes = FALSE) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Distribution of cdn_way as predicted by EUI",
       x = "EUI",
       y = "Percentage",
       fill = "Is there a Canadian Way?") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0),
        axis.text.x = element_text(angle = 17))




p4 + scale_fill_manual(values = c("#0072B2","#CC79A7", "#E69F00", "#F0E442",  "#D55E00")) 

```

Is what we see here statistically significant? 

```{r}

# Create a contingency table
contingency_table <- df_plot %>%
  count(eui_bin, cdn_way) %>%
  spread(cdn_way, n, fill = 0)

# Perform the chi-square test
chi_square_test <- chisq.test(as.matrix(contingency_table[,-1]))

# Print the results
print(chi_square_test)

```
::: {.callout-tip}
## The lowdown
Significant at the *p* ≤ 0.05<sup>*</sup> level.
:::

## EUI and SLAI

```{r fig-slai-eui, warning=FALSE}

df_eui <- import("data/df_canada_6.rds") %>% 
  tibble() %>% 
  filter(!region=="other") %>% 
  mutate(
    eui = as.integer(eui),
    eui_bin = case_when(
    eui == 0 ~ "Monolingual (0)",
    eui >= 1 & eui <= 2 ~ "Some non-English (1-2)",
    eui >= 3 & eui <= 15 ~ "Considerable non-English (3-15)",
    TRUE ~ NA_character_
  ),
  eui_bin = factor(eui_bin, levels = c("Monolingual (0)", "Some non-English (1-2)", "Considerable non-English (3-15)"))
  )

df_plot <- df_eui %>% 
  select(eui_bin, slai) %>% 
  na.omit()

# Create the plot
p6 <- df_plot %>% 
  ggplot(aes(x = eui_bin, y = slai, fill = eui_bin)) +
   geom_boxplot(aes(fill = eui_bin), width = 0.7,
                show.legend = F) +
   labs(title = "Distribution of SLAI by EUI",
             x = "EUI",
             y = "SLAI") +
   theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0))

p6 + scale_fill_manual(values = c("#D55E00", "#0072B2","#CC79A7", "#E69F00", "#F0E442")[c(2,4,5)])

```

::: {.callout-tip}
## The lowdown
This distribution is not at all significant.
:::
